Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2025-01-23 LICS_2025/monoidal_e_graphs/sections/combinatorial_semantics.tex
cleanup prose, make fit in 12 pages

In this section,  we recall the fundamental definitions and results of string diagram rewrite theory for symmetric monoidal categories,  building to the recapitulation of the correspondence between string diagram rewriting and an appropriate notion of double pushout (DPO) rewriting of certain hypergraphs,  as established in~\cite{bonchi_string_2022-1, bonchi_string_2022-2}.

@@

In particular, this means that any morphism in $\HypI{\Sigma}$ is of the form $n \xrightarrow{f} \mathcal{G} \xleftarrow{g} m$,
where $n,\;m$ are discrete ordered hypergraphs and $\mathcal{G}$ is an arbitrary hypergraph.  The morphisms $f$ and $g$ into $\mathcal{G}$ will specify which vertices are to be considered as inputs and outputs,  respectively.
The ordering of the interface hypergraphs is the way to differentiate between the interpretations of morphisms $f \otimes g$ and $g \otimes f$.

@@

\begin{definition}[Convex DPOI rewriting]
\label{def:convex_dpo}
Let $\mathfrak{R}$ be a set of DPOI rewrite rules. 
Then, given $\mathcal G \xleftarrow{} n+m$ and $\mathcal H \xleftarrow{} n + m$ in $\catname{Hyp(\Sigma)}$, $\mathcal G$ rewrites convexly into $\mathcal H$ with interface $n + m$ --- notation $(\mathcal G \xleftarrow{} n + m ) \Rrightarrow_{\mathfrak{R}}  (\mathcal H \xleftarrow{} n + m )$ --- if there exist rule $\mathcal L \xleftarrow{} i + j \xrightarrow{} \mathcal R$ in $\mathfrak{R}$ and object $\mathcal{L}^{\bot}$ and cospan arrows $i+j \xrightarrow{} \mathcal{L}^{\bot} \xleftarrow{} n+m$ such that the DPOI diagram of Definition \ref{def:dpoi} commutes and its marked squares are pushouts 
and the following conditions hold
\begin{itemize}
    \item $f : L \to \mathcal G$ is a convex match;
    \item $i + j \to \mathcal{L}^{\bot} \to \mathcal G$ is a boundary complement in the leftmost pushout.
\end{itemize}
\end{definition}
%By requiring convex matching and boundary complements we ensure that the original cospans above are monogamous, e.g., $n \xrightarrow{} G \xleftarrow{} m$.
It is then proved in~\cite{bonchi_string_2022-2} that this notion of DPOI rewriting is sound and complete for $\catname{S}(\Sigma, \mathcal{E})$: for any morphisms $f$ and $g$ in $\catname{S}(\Sigma)$,  we have that $f = g$ modulo $\mathcal{E}$ if and only if there exist a sequence of DPO rewrites (induced by $\mathcal{E}$) between the representations of $f$ and $g$ as (cospans of) hypergraphs.
As SMC equations get absorbed by the representation, it makes this approach computationally appealing.
% In particular,  Theorem \ref{thm:prop-equiv} means that the structural SMC equations are factored out in the representation of $f$ and $g$,  and the DPO rewrites necessary are only those induced by $\mathcal{E}$. 
%\begin{theorem}[Theorem 35~\cite{bonchi_string_2022-2}]
%For any $f,g \in \mathbf{SMT}(\Sigma, \mathcal{E})$,  we have that 
%\[
%	f = g \iff \llangle \lceil f \rceil \rrangle \Rrightarrow^*_{\mathfrak{R}} \llangle \lceil g \rceil \rrangle~, 
%\]
%where $\mathfrak{R} = \llangle \lceil \mathcal{E} \rceil \rrangle$ and $ \Rrightarrow^*_{\mathfrak{R}}$ denotes the existence of a sequence of convex rewrites. 
%\end{theorem}

>>>>

In this section,  we recall the fundamental definitions and results of string diagram rewrite theory for symmetric monoidal categories,  recapitulating the correspondence between string diagram rewriting and an appropriate notion of double pushout (DPO) rewriting of certain hypergraphs,  as established in~\cite{bonchi_string_2022-1, bonchi_string_2022-2}.

@@

In particular, this means that any morphism in $\HypI{\Sigma}$ is of the form $n \xrightarrow{f} \mathcal{G} \xleftarrow{g} m$,
where $n,\;m$ are discrete ordered hypergraphs and $\mathcal{G}$ is an arbitrary hypergraph.  The morphisms $f$ and $g$ into $\mathcal{G}$ will specify which vertices are to be considered as inputs and outputs,  respectively.
The ordering of the interface associated to a hypergraph is used to differentiate between the interpretations of morphisms $f \otimes g$ and $g \otimes f$.

@@

\begin{definition}[Convex DPOI rewriting]
\label{def:convex_dpo}
Let $\mathfrak{R}$ be a set of DPOI rewrite rules. 
Then, given $\mathcal G \xleftarrow{} n+m$ and $\mathcal H \xleftarrow{} n + m$ in $\catname{Hyp(\Sigma)}$, $\mathcal G$ rewrites convexly into $\mathcal H$ with interface $n + m$ --- notation $(\mathcal G \xleftarrow{} n + m ) \Rrightarrow_{\mathfrak{R}}  (\mathcal H \xleftarrow{} n + m )$ --- if there exist rule $\mathcal L \xleftarrow{} i + j \xrightarrow{} \mathcal R$ in $\mathfrak{R}$ and object $\mathcal{L}^{\bot}$ and cospan arrows $i+j \xrightarrow{} \mathcal{L}^{\bot} \xleftarrow{} n+m$ such that the DPOI diagram of Definition \ref{def:dpoi} commutes and its marked squares are pushouts 
and the following conditions hold
\begin{itemize}
    \item $f : L \to \mathcal G$ is a convex match;
    \item $i + j \to \mathcal{L}^{\bot} \to \mathcal G$ is a boundary complement in the leftmost pushout.
\end{itemize}
\end{definition}
%By requiring convex matching and boundary complements we ensure that the original cospans above are monogamous, e.g., $n \xrightarrow{} G \xleftarrow{} m$.
It is then proved in~\cite{bonchi_string_2022-2} that this notion of DPOI rewriting is sound and complete for $\catname{S}(\Sigma, \mathcal{E})$: for any morphisms $f$ and $g$ in $\catname{S}(\Sigma)$,  we have that $f = g$ modulo $\mathcal{E}$ if and only if there exist a sequence of DPO rewrites (induced by $\mathcal{E}$) between the representations of $f$ and $g$ as (cospans of) hypergraphs.
As SMC equations get absorbed by the representation, it makes this approach computationally appealing as they do not need to be represented explicitly.
% In particular,  Theorem \ref{thm:prop-equiv} means that the structural SMC equations are factored out in the representation of $f$ and $g$,  and the DPO rewrites necessary are only those induced by $\mathcal{E}$. 
%\begin{theorem}[Theorem 35~\cite{bonchi_string_2022-2}]
%For any $f,g \in \mathbf{SMT}(\Sigma, \mathcal{E})$,  we have that 
%\[
%	f = g \iff \llangle \lceil f \rceil \rrangle \Rrightarrow^*_{\mathfrak{R}} \llangle \lceil g \rceil \rrangle~, 
%\]
%where $\mathfrak{R} = \llangle \lceil \mathcal{E} \rceil \rrangle$ and $ \Rrightarrow^*_{\mathfrak{R}}$ denotes the existence of a sequence of convex rewrites. 
%\end{theorem}

### 2025-01-23 LICS_2025/monoidal_e_graphs/sections/e-hypergraphs.tex
cleanup prose, make fit in 12 pages

The category of e-hypergraphs has coproducts,  given by the disjoint union of e-hypergraphs,  and an initial object given by the empty e-hypergraph.  
A concrete description of the pushout of two morphisms in this category is given in the Appendix section~\ref{sec:appendix:pushout}.
Generally,  the pushout of two e-hypergraph homomorphisms need not exist,  but we prove that it does in all scenarios that we are interested in, most importantly the pushout for the composition of two cospans with discrete feet exists.

@@

When modelling enriched string diagrams using e-hypergraphs,  we require keeping track of more than simply the ultimate external inputs and outputs of the diagram: each internal e-box has its own inputs and outputs,  which we must also keep track of.  
Thus,  we introduce an \textit{extended} cospan construction.

@@

Consider two extended cospans below 
\[
\scalebox{0.75}{
    \tikzfig{../figures/combinatorial_semantics/isomorphic_e_cospans}
}
\]
and consider a relation 
\[
R = \{x R y \text{ if } f_{int}(x) \consistency f_{int}(y) \text{ for } x,y \in n'\}
\]
and let $S$ be its reflexive closure.
The latter partitions $n'$ into non-empty subsets $\{p_{j}\}_{j=1}^{k}$. 
We get an analogous partition for $m'$, $n''$, $m''$.
Two extended cospans are isomorphic if there exist isomorphisms $\alpha$, $\beta$ and $\gamma$ making the above diagram commute and such that $\alpha$ and $\gamma$ preserve order within each $p_i$.
\end{definition}
Intuitively, such construction suggests that the position of the image of $n$ within $n'$ relative to the strict internal interface does not matter; and that when considering individual components of hierarchical edges with their respective portions of strict internal input and output interfaces as plain cospans of hypergraphs (by only considering top-level vertices and edges of these components) they should be isomorphic cospans of hypergraphs in the sense from~\cite{bonchi_string_2022-1}.
An example of an extended cospan can be seen in Figure~\ref{fig:e-cospan-example}.
The partition of the interfaces is given with colours.
The labels suggest how $f_{int}, f_{ext}$ (respectively, $g_{int}, g_{ext}$) act on vertices.
The ordering of vertices in the interfaces is from left to right.
Examples of isomorphic and non-isomorphic extended cospans can be found in Appendix section~\ref{sec:appendix:iso}.

@@

Composition of two morphisms 
\begin{align*}
	n \xrightarrow{f_{ext}} n' \xrightarrow{f_{int}} &\mathcal{F} \xleftarrow{f'_{int}} m' \xleftarrow{f'_{ext}} m\\
	m \xrightarrow{g_{ext}} m'\!' \xrightarrow{g_{int}} &\mathcal{G} \xleftarrow{g'_{int}} k' \xleftarrow{g'_{ext}} k
\end{align*} is computed in two stages.
First, $\mathcal{H}$ is computed as the result of the pushout square shown below; 
% https://q.uiver.app/#q=WzAsMTAsWzAsMCwiZXh0KFgpIl0sWzEsMSwiWCJdLFsyLDIsIkEiXSxbMywwLCJZIl0sWzQsMCwiZXh0KFkpIl0sWzUsMCwiWSciXSxbNiwyLCJCIl0sWzcsMSwiWiJdLFs4LDAsImV4dChaKSJdLFs0LDMsIl57XFx1bGNvcm5lcn1BK197Zl97ZXh0fTtmLGdfe2V4dH07Z31CIl0sWzAsMV0sWzEsMl0sWzMsMiwiZiJdLFs0LDMsImZfe2V4dH0iXSxbNCw1LCJnX3tleHR9IiwyXSxbNSw2LCJnIiwyXSxbNyw2XSxbOCw3XSxbMiw5XSxbNiw5XV0=
\[
\trimbox{0cm 0cm 0cm 0.75cm}{
\adjustbox{scale=0.9,center}
{\tikzfig{../figures/combinatorial_semantics/pushout_e_cospans}}
}
\]
then,  the result of composition is defined as follows:
\[
n \xrightarrow{f_{ext};\iota_1} n' + (m'' \setminus m) \xrightarrow{h_{1}} \mathcal{H} \xleftarrow{h_2} k' + (m' \setminus m) \xleftarrow{g'_{ext};\iota_1} k
\]
where $h_i$ are defined as follows,  using $|$ to denote the restriction of a function on a discrete e-hypergraph. 
\[
    h_1 = [ f_{int};p_1, ~(g_{int};p_2)|_{m'' \setminus m} ]
\;
    h_2 = [ g'_{int};p_2, ~(f'_{int};p_1)|_{m' \setminus m} ]
\]
% \[
% \begin{tikzcd}
% Y_{ext} \arrow[d] & 0 \arrow[l] \arrow[d]    \\
% Y'          & Y' \setminus Y_{ext} \arrow[l]
% \end{tikzcd}
% \]
% and then the disjoint union is just a coproduct. 
% where $g'$ is the restriction of $g$ to ${Y' \setminus Y_{ext}}$ and $f'$ the restriction of $f$ to $Y \setminus Y_{ext}$. 
The identity of composition is given by the obvious extended cospan of identities. 
$\Ecospans$ inherits a symmetric monoidal structure from the coproduct (and initial object) structure of $\catname{EHyp({\Sigma}})$,  analogously to Definition~\ref{def:cspd}.

@@

Because extended cospans have a more general notion of interface,  including \textit{internal} vertices,  DPOI rewriting as presented in Section \ref{sec:combinatorial-semantics} needs some adjustments. This is the purpose of this section of the paper.

We do not expect internal interfaces to be preserved during rewriting: for example,  when the semilattice equations are considered as rewrites. 
Thus,  we wish a rewrite rule to be a pair of \textit{extended} cospans of e-hypergraphs with matching \textit{external} (but not necessarily internal) interfaces,  as follows. 
\[
    n \xrightarrow{} n' \xrightarrow{} \mathcal{L} \xleftarrow{} m' \xleftarrow{} m
\qquad
    n \xrightarrow{} n'' \xrightarrow{} \mathcal{R} \xleftarrow{} m'' \xleftarrow{} m
\] 
Analogous to Section \ref{sec:combinatorial-semantics}, observing that the following extended cospans express the same data
\[
    0 \xrightarrow{} 0 \xrightarrow{} \mathcal{L} \xleftarrow{} n'' + m'' \xleftarrow{} n + m
\;
    0 \xrightarrow{} 0 \xrightarrow{} \mathcal{R} \xleftarrow{} n' + m' \xleftarrow{} n + m
\] allows us to encode rewrite rules as extended cospans of the following form
\[
\mathcal{L} \xleftarrow{} n' + m' \xleftarrow{} n + m \xrightarrow{} n'' + m'' \xrightarrow{} \mathcal{R}
\] which will fit into the DPO formalism. We make this identification throughout,  without further mention. 
%We will futher omit the arrows $0 \to 0 \to \mathcal{L}$.

@@

First, note that conditions (6) and (7) are disjoint.
Intuitively the above definition means that when the occurrence of $\mathcal{L}$ within $\mathcal{G}$ is top-level (premise for condition (6)) then $\mathcal{L}$'s external interfaces become the part of $\mathcal{L}^{\bot}$'s external interfaces (as in the case of Definition~\ref{def:dpoi}) and when the occurrence is nested (premise of condition (7)), those interfaces become the part of $\mathcal{L}^{\bot}$'s strict internal interfaces: one can think about the latter as applying Definition~\ref{def:dpoi} in a nested context.
In both cases when removing the occurrence of $\mathcal{L}$ from $\mathcal{G}$, $\mathcal{L}$'s strict internal interfaces are removed from strict internal interfaces of $\mathcal{G}$ to form internal interfaces of the corresponding MDA cospans above.
The morphisms are then constructed correspondingly to require that inputs (respectively, outputs) of $\mathcal{R}$ are glued to the outputs (respectively, inputs) of $\mathcal{L}^{\bot}$.

@@

Note that,  in the above definition,  the extended cospan 
    \[
    \vspace{-1mm}
    n \xrightarrow{} n' \setminus (i' \setminus i) + j \xrightarrow{[g_1,c_2]} \mathcal{L}^{\bot} \xleftarrow{[g_2,c_1]} k' \setminus (j' \setminus j) + i \xleftarrow{} k
    \vspace{-0.5mm}
    \]
 is not necessarily well-typed.
This is because its input internal interface may contain vertices that previously were a part of the output internal interface and vice versa.
Examples and non-examples of a boundary complement can be found in Appendix section~\ref{sec:appendix:iso}.
% \begin{remark}
% The boundary complement conditions, in particular, prohibit finding a match for $\mathcal L$ in $\mathcal G$,  below.  
% \[
% 	\mathcal{L} = \scalebox{0.8}{\tikzfig{../figures/combinatorial_semantics/f_times_g}} \qquad \mathcal{G} = \scalebox{0.6}{\tikzfig{../figures/combinatorial_semantics/f_plus_g_inline}}
% \]
% \end{remark}

\begin{proposition}
\label{prop:boundary_unique}
    The boundary complement in~\ref{def:boundary_new} when exists is unique.
\end{proposition}

@@

\begin{definition}[Convex EDPOI rewriting]
\label{def:dpoi-e}
Given an extended span of morphisms 
\[
    \mathcal{G} \xleftarrow{} n' + k' \xleftarrow{} n + k \xrightarrow{} n'' + k'' \xrightarrow{} \mathcal{H}
\]
in $\catname{EHyp(\Sigma)}$, we say $\mathcal{G}$ \textit{rewrites (convexly) to} $\mathcal{H}$ (\textit{with external interface} $n + k$ and \textit{taking internal interface} $n'+k'$ \textit{to} $n'' + k''$) --- denoted by $\mathcal{G} \Rrightarrow \mathcal H$  --- \textit{via a rewrite rule} 
\[
    \mathcal{L} \xleftarrow{} i' + j' \xleftarrow{} i + j \xrightarrow{} i'' + j'' \xrightarrow{} \mathcal{R}
\] 
if there exists an object $\mathcal{C}$ and morphisms which complete the following commutative diagram 
\[
 \scalebox{0.75}{
    \tikzfig{../figures/combinatorial_semantics/DPOI_square}
 }
\]
such that the two marked squares are pushouts and the following conditions hold:
    \begin{enumerate}
        \label{dpoi-e:assumptions}
       % \item $m : \mathcal{L} \to \mathcal{G}$ is a convex match;
        %\item $m(L)$ is a down-closed e-hypergraph;
        \item $i + j \to \mathcal{C} \to \mathcal{G}$ is a boundary complement;
	\item the internal interfaces of $\mathcal H$ are such that:
	\[
		n'' = n' \setminus (i' \setminus i) + (i'' \setminus i) \qquad k'' = k' \setminus (j' \setminus j) +  (j'' \setminus j)\\
	\]
	\item the map $f_1 = [g_1,h_1]: n'' \to \mathcal H$ in the diagram above consists of $g_1$ as defined in Definition \ref{def:boundary_new} of boundary complement,  and $h_1$ which is the restriction of the composite $i'' \to \mathcal R \to \mathcal H$ to $i'' \setminus i$,  and similarly for the map $f_2: k'' \to \mathcal H$.  
    \end{enumerate}
% The internal interfaces of the resulting e-hypergraph $\mathcal{H}$ from the cospan
%  \[
%     \mathcal{H} \xleftarrow{} n'' + k'' \xleftarrow{} n + k
% \]
\end{definition}
Given  a set $\mathfrak{R}$ of EDPOI rewrite rules and e-hypergraphs with extended interfaces $\mathcal G$ and $\mathcal H$,  we write $\mathcal G \Rrightarrow_\mathfrak{R} \mathcal H$ if there exists a EDPOI rewrite in $\mathfrak R$ such that via it $\mathcal G$ rewrites to $\mathcal H$,  and we write $\mathcal G \Rrightarrow^*_\mathfrak{R} \mathcal H$ for reflexive, symmetric, and transitive closure of $\Rrightarrow{}$.
An example of rewriting of extended cospans can be found in Appendix section~\ref{sec:appendix:iso}.
% Together, the conditions ensure the result of a rewrite is a well-typed MDA e-hypergraph.

>>>>

The category of e-hypergraphs has coproducts,  given by the disjoint union of e-hypergraphs,  and an initial object given by the empty e-hypergraph.  
A concrete description of the pushout of two morphisms in this category is given in Appendix~\ref{sec:appendix:pushout}.
Generally,  the pushout of two e-hypergraph homomorphisms need not exist,  but we prove that it does in all scenarios that we are interested in; most importantly, the pushout for the composition of two cospans with discrete feet always exists.

@@

When modelling enriched string diagrams using e-hypergraphs,  we must keep track of more than simply the ultimate external inputs and outputs of the diagram: each internal e-box has its own inputs and outputs,  which also must be tracked.  
Thus,  we introduce an \textit{extended} cospan construction.

@@

Consider two extended cospans below 
\[
\scalebox{0.75}{
    \tikzfig{../figures/combinatorial_semantics/isomorphic_e_cospans}
}
\]
and consider a relation 
\[
R = \{x R y \text{ if } f_{int}(x) \consistency f_{int}(y) \text{ for } x,y \in n'\}
\]
and let $S$ be its reflexive closure.
The latter partitions $n'$ into non-empty subsets $\{p_{j}\}_{j=1}^{k}$. 
We get an analogous partition for $m'$, $n''$, $m''$.
Two extended cospans are isomorphic if there exist isomorphisms $\alpha$, $\beta$ and $\gamma$ making the above diagram commute and such that $\alpha$ and $\gamma$ preserve order within each $p_i$.
\end{definition}
Intuitively, such construction suggests that the position of the image of $n$ within $n'$ relative to the strict internal interface does not matter; and that when considering individual components of hierarchical edges with their respective portions of strict internal input and output interfaces as plain cospans of hypergraphs (by only considering top-level vertices and edges of these components) they should be isomorphic cospans of hypergraphs in the sense from~\cite{bonchi_string_2022-1}.
An example of an extended cospan can be seen in Figure~\ref{fig:e-cospan-example}.
The partition of the interfaces is given with colours.
The labels suggest how $f_{int}, f_{ext}$ (respectively, $g_{int}, g_{ext}$) act on vertices, with the ordering of vertices in the interfaces given left to right.
Examples of isomorphic and non-isomorphic extended cospans can be found in Appendix~\ref{sec:appendix:iso}.

@@

Composition of two morphisms 
\begin{align*}
	n \xrightarrow{f_{ext}} n' \xrightarrow{f_{int}} &\mathcal{F} \xleftarrow{f'_{int}} m' \xleftarrow{f'_{ext}} m\\
	m \xrightarrow{g_{ext}} m'\!' \xrightarrow{g_{int}} &\mathcal{G} \xleftarrow{g'_{int}} k' \xleftarrow{g'_{ext}} k
\end{align*} is computed in two stages.
First, $\mathcal{H}$ is computed as the result of the pushout square shown below: 
% https://q.uiver.app/#q=WzAsMTAsWzAsMCwiZXh0KFgpIl0sWzEsMSwiWCJdLFsyLDIsIkEiXSxbMywwLCJZIl0sWzQsMCwiZXh0KFkpIl0sWzUsMCwiWSciXSxbNiwyLCJCIl0sWzcsMSwiWiJdLFs4LDAsImV4dChaKSJdLFs0LDMsIl57XFx1bGNvcm5lcn1BK197Zl97ZXh0fTtmLGdfe2V4dH07Z31CIl0sWzAsMV0sWzEsMl0sWzMsMiwiZiJdLFs0LDMsImZfe2V4dH0iXSxbNCw1LCJnX3tleHR9IiwyXSxbNSw2LCJnIiwyXSxbNyw2XSxbOCw3XSxbMiw5XSxbNiw5XV0=
\[
\trimbox{0cm 0cm 0cm 0.75cm}{
\adjustbox{scale=0.9,center}
{\tikzfig{../figures/combinatorial_semantics/pushout_e_cospans}}
}
\]
then,  the result of composition is defined as follows:
\[
n \xrightarrow{f_{ext};\iota_1} n' + (m'' \setminus m) \xrightarrow{h_{1}} \mathcal{H} \xleftarrow{h_2} k' + (m' \setminus m) \xleftarrow{g'_{ext};\iota_1} k
\]
where $h_i$ are defined as follows,  using $|$ to denote the restriction of a function on a discrete e-hypergraph. 
\[
    h_1 = [ f_{int};p_1, ~(g_{int};p_2)|_{m'' \setminus m} ]
\;
    h_2 = [ g'_{int};p_2, ~(f'_{int};p_1)|_{m' \setminus m} ]
\]
% \[
% \begin{tikzcd}
% Y_{ext} \arrow[d] & 0 \arrow[l] \arrow[d]    \\
% Y'          & Y' \setminus Y_{ext} \arrow[l]
% \end{tikzcd}
% \]
% and then the disjoint union is just a coproduct. 
% where $g'$ is the restriction of $g$ to ${Y' \setminus Y_{ext}}$ and $f'$ the restriction of $f$ to $Y \setminus Y_{ext}$. 
The identity of composition is given by the obvious extended cospan of identities. 
$\Ecospans$ inherits a symmetric monoidal structure from the coproduct (and initial object) structure of $\catname{EHyp({\Sigma}})$,  analogously to Definition~\ref{def:cspd}.

@@

Because extended cospans have a more general notion of interface,  including \textit{internal} vertices,  DPOI rewriting as presented in Section \ref{sec:combinatorial-semantics} needs some adjustments.

We do not expect internal interfaces to be preserved during rewriting: for example,  when the semilattice equations are considered as rewrites. 
Thus,  we wish for a rewrite rule to be a pair of \textit{extended} cospans of e-hypergraphs with matching \textit{external} (but not necessarily internal) interfaces,  as follows: 
\[
    n \xrightarrow{} n' \xrightarrow{} \mathcal{L} \xleftarrow{} m' \xleftarrow{} m,
\qquad
    n \xrightarrow{} n'' \xrightarrow{} \mathcal{R} \xleftarrow{} m'' \xleftarrow{} m.
\] 
Analogously to Section \ref{sec:combinatorial-semantics}, observing that the following extended cospans express the same data:
\[
    0 \xrightarrow{} 0 \xrightarrow{} \mathcal{L} \xleftarrow{} n'' + m'' \xleftarrow{} n + m
\;
    0 \xrightarrow{} 0 \xrightarrow{} \mathcal{R} \xleftarrow{} n' + m' \xleftarrow{} n + m
\] allows us to encode rewrite rules as extended cospans of the following form:
\[
\mathcal{L} \xleftarrow{} n' + m' \xleftarrow{} n + m \xrightarrow{} n'' + m'' \xrightarrow{} \mathcal{R},
\] which will fit into the DPO formalism. We make this identification throughout,  without further mention. 
%We will futher omit the arrows $0 \to 0 \to \mathcal{L}$.

@@

First, note that conditions (6) and (7) are disjoint.
Intuitively, the above definition means that when the occurrence of $\mathcal{L}$ within $\mathcal{G}$ is top-level (premise for condition (6)) then $\mathcal{L}$'s external interfaces become the part of $\mathcal{L}^{\bot}$'s external interfaces (as in the case of Definition~\ref{def:dpoi}), and when the occurrence is nested (premise of condition (7)), those interfaces become the part of $\mathcal{L}^{\bot}$'s strict internal interfaces: one can think about the latter as applying Definition~\ref{def:dpoi} in a nested context.
In both cases, when removing the occurrence of $\mathcal{L}$ from $\mathcal{G}$, $\mathcal{L}$'s strict internal interfaces are removed from strict internal interfaces of $\mathcal{G}$ to form internal interfaces of the corresponding MDA cospans above.
The morphisms are then constructed correspondingly to require that inputs (respectively, outputs) of $\mathcal{R}$ are glued to the outputs (respectively, inputs) of $\mathcal{L}^{\bot}$.

@@

Note that,  in the above definition,  the extended cospan 
    \[
    \vspace{-1mm}
    n \xrightarrow{} n' \setminus (i' \setminus i) + j \xrightarrow{[g_1,c_2]} \mathcal{L}^{\bot} \xleftarrow{[g_2,c_1]} k' \setminus (j' \setminus j) + i \xleftarrow{} k
    \vspace{-0.5mm}
    \]
 is not necessarily well-typed.
This is because its input internal interface may contain vertices that previously were a part of the output internal interface and vice versa.
Examples and non-examples of a boundary complement can be found in Appendix~\ref{sec:appendix:iso}.
% \begin{remark}
% The boundary complement conditions, in particular, prohibit finding a match for $\mathcal L$ in $\mathcal G$,  below.  
% \[
% 	\mathcal{L} = \scalebox{0.8}{\tikzfig{../figures/combinatorial_semantics/f_times_g}} \qquad \mathcal{G} = \scalebox{0.6}{\tikzfig{../figures/combinatorial_semantics/f_plus_g_inline}}
% \]
% \end{remark}

\begin{proposition}
\label{prop:boundary_unique}
    The boundary complement in~\ref{def:boundary_new} when it exists is unique.
\end{proposition}

@@

\begin{definition}[Convex EDPOI rewriting]
\label{def:dpoi-e}
Given an extended span of morphisms 
\[
    \mathcal{G} \xleftarrow{} n' + k' \xleftarrow{} n + k \xrightarrow{} n'' + k'' \xrightarrow{} \mathcal{H}
\]
in $\catname{EHyp(\Sigma)}$, we say $\mathcal{G}$ \textit{rewrites (convexly) to} $\mathcal{H}$ (\textit{with external interface} $n + k$ and \textit{taking internal interface} $n'+k'$ \textit{to} $n'' + k''$) --- denoted by $\mathcal{G} \Rrightarrow \mathcal H$  --- \textit{via a rewrite rule} 
\[
    \mathcal{L} \xleftarrow{} i' + j' \xleftarrow{} i + j \xrightarrow{} i'' + j'' \xrightarrow{} \mathcal{R}
\] 
if there exists an object $\mathcal{C}$ and morphisms which complete the following commutative diagram 
\[
 \scalebox{0.75}{
    \tikzfig{../figures/combinatorial_semantics/DPOI_square}
 }
\]
such that the two marked squares are pushouts and the following conditions hold:
    \begin{enumerate}
        \label{dpoi-e:assumptions}
       % \item $m : \mathcal{L} \to \mathcal{G}$ is a convex match;
        %\item $m(L)$ is a down-closed e-hypergraph;
        \item $i + j \to \mathcal{C} \to \mathcal{G}$ is a boundary complement;
	\item the internal interfaces of $\mathcal H$ are such that:
	\[
		n'' = n' \setminus (i' \setminus i) + (i'' \setminus i) \qquad k'' = k' \setminus (j' \setminus j) +  (j'' \setminus j)\\
	\]
	\item the map $f_1 = [g_1,h_1]: n'' \to \mathcal H$ in the diagram above consists of $g_1$ as defined in Definition \ref{def:boundary_new} of boundary complement,  and $h_1$ which is the restriction of the composite $i'' \to \mathcal R \to \mathcal H$ to $i'' \setminus i$,  and similarly for the map $f_2: k'' \to \mathcal H$.  
    \end{enumerate}
% The internal interfaces of the resulting e-hypergraph $\mathcal{H}$ from the cospan
%  \[
%     \mathcal{H} \xleftarrow{} n'' + k'' \xleftarrow{} n + k
% \]
\end{definition}
Given  a set $\mathfrak{R}$ of EDPOI rewrite rules and e-hypergraphs with extended interfaces $\mathcal G$ and $\mathcal H$,  we write $\mathcal G \Rrightarrow_\mathfrak{R} \mathcal H$ if there exists a EDPOI rewrite in $\mathfrak R$ such that via it $\mathcal G$ rewrites to $\mathcal H$,  and we write $\mathcal G \Rrightarrow^*_\mathfrak{R} \mathcal H$ for reflexive, symmetric, and transitive closure of $\Rrightarrow{}$.
An example of rewriting of extended cospans can be found in Appendix~\ref{sec:appendix:iso}.
% Together, the conditions ensure the result of a rewrite is a well-typed MDA e-hypergraph.
